SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-01 method.tex
Minor change
Updated results section & added graphs

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2 cache
demand accesses but could be directed to the L2 or last-level cache. No L1
data level prefetching is done. The LRU replacement policy is used on all levels
of cache hierarchies. Branch prediction is done using the perceptron branch
predictor~\cite{Perc_Branch}.  ThE page size is 4KB.  Champsim operates
all the prefetchers strictly in the physical address space.

@@

%This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the that SPEC 2017 benchmark suite~\cite{SPEC2017}.  We use all the 20
workloads available in the SPEC 2017 suite.  Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

@@

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|m{4.8cm}|}
    \hline
        \textbf{Field} &
        \textbf{Bits} &
        \textbf{Comment} \\
    \hline
         Valid & 1 & Indicates a valid entry in the table\\
         Tag & 6 & Identifier for the entry in the table\\
         Useful & 1 & To show if the given entry led to a useful demand fetch\\
         Perc Decision & 1 & Prefetched vs Not-prefetched \\
    \hline
        PC & 12 & \multirow{5}{4.8cm}{Meta-data required for perceptron training}\\
        Address & 24 & \\
        Delta & 7 & \\
        Curr Signature & 12 & \\
        Confidence & 7 & \\
    \hline
        \multicolumn{3}{|c|}{\textbf{Total 84 bits}}\\
    \hline
    \end{tabular}
    \caption{Meta-data Stored in Prefetch Table}
    \label{tab:PTable_metadata}
\end{table}

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
        \multirow{5}{2.2cm}{Signature Table} &      & Valid (1 bit)        &             \\
                                             &      & Tag (16 bits)        &             \\
                                             & 256  & Last Offset (6 bits) & 11008 bits  \\  
                                             &      & Signature (12 bits)  &             \\
                                             &      & LRU (6 bits)         &             \\
    \hline
        \multirow{3}{2.2cm}{~~Pattern Table} &      & $C_{sig}$ (4bits)      &               \\
                                             & 512  & $C_{delta}$ (4*4 bits) & 24576 bits    \\
                                             &      & Delta (4*7 bits)       &               \\
    \hline
        \multirow{3}{1.5cm}{Perceptron\newline}     & 4096*8    &            &              \\
        \multirow{2}{0.9cm}{Weights}                 & 2048*x    & 5 bits     & 164480 bits  \\
                                                    & 128*1     &            &              \\
    \hline
        Prefetch Table\footnotemark[1]             & 1024      & 84 bits    & 86016 bits   \\
    \hline
        Reject Table\footnotemark[2]               & 1024      & 83 bits    & 84992 bits   \\
    \hline
        \multirow{4}{1cm}{Global\newline\newline}   & \multirow{4}{0.2cm}{8} & Signature (12 bits)  & \multirow{4}{1.1cm}{264 bits} \\
        \multirow{3}{1.1cm}{History\newline}        &                        & Confidence (8 bits)  &                               \\
        \multirow{2}{1.2cm}{Register}               &                        & Last Offset (6 bits) &                               \\
                                                    &                        & Delta (7 bits)       &                               \\
    \hline
        Accuracy        & 1     & C$_{total}$       & 10 bits   \\
        Counters        & 1     & C$_{useful}$      & 10 bits   \\
    \hline
        \multirow{3}{1.5cm}{Global PC\newline}      &       & $PC_1$ (12 bits)      &           \\
        \multirow{2}{1.5cm}{~Trackers}              & 3     & $PC_2$ (12 bits)      & 36 bits   \\
                                                    &       & $PC_3$ (12 bits)      &           \\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 371392 bits = 45.33 KB}}\\
    \hline
    \end{tabular}
    \caption{SPP-Perc Storage Overhead}
    \label{tab:PPF_overhead}
\end{table}

\footnotetext[1]{Components of Prefetch Table can be found in Table \ref{tab:PTable_metadata}}
\footnotetext[2]{the Reject Table does not need to maintain the useful bit as that only applies for prefetches that ultimately made through}

@@

>>>>

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2 cache
demand accesses but could be directed to the L2 or last-level cache. No L1
data level prefetching is done. The LRU replacement policy is used on all levels
of cache hierarchies. Branch prediction is done using the perceptron branch
predictor~\cite{Perc_Branch}.  The page size is 4KB.  Champsim operates
all the prefetchers strictly in the physical address space.

@@

%This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the SPEC 2017 benchmark suite~\cite{SPEC2017}.  We use all the 20
workloads available in the SPEC 2017 suite.  Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

@@

\begin{table}[t]
    \centering
    \begin{tabular}{|c|c|m{4.8cm}|}
    \hline
        \textbf{Field} &
        \textbf{Bits} &
        \textbf{Comment} \\
    \hline
         Valid & 1 & Indicates a valid entry in the table\\
         Tag & 6 & Identifier for the entry in the table\\
         Useful & 1 & To show if the given entry led to a useful demand fetch\\
         Perc Decision & 1 & Prefetched vs Not-prefetched \\
    \hline
        PC & 12 & \multirow{5}{4.8cm}{Meta-data required for perceptron training}\\
        Address & 24 & \\
        Delta & 7 & \\
        Curr Signature & 12 & \\
        Confidence & 7 & \\
    \hline
        \multicolumn{3}{|c|}{\textbf{Total 84 bits}}\\
    \hline
    \end{tabular}
    \caption{Meta-data Stored in Prefetch Table}
    \label{tab:PTable_metadata}
\end{table}

@@

\begin{table}[h]
\begin{adjustwidth}{-0.5cm}{}
    \centering
    \begin{tabular}{|c|c|c|c|}
    \hline
        \textbf{Structure} &
        \textbf{Entry} &
        \textbf{Components} &
        \textbf{Total} \\
    \hline
        \multirow{5}{1.5cm}{Signature Table} &      & Valid (1 bit)        &             \\
                                             &      & Tag (16 bits)        &             \\
                                             & 256  & Last Offset (6 bits) & 11008 bits  \\  
                                             &      & Signature (12 bits)  &             \\
                                             &      & LRU (6 bits)         &             \\
    \hline
        \multirow{3}{2cm}{~~Pattern Table} &      & $C_{sig}$ (4bits)      &               \\
                                             & 512  & $C_{delta}$ (4*4 bits) & 24576 bits    \\
                                             &      & Delta (4*7 bits)       &               \\
    \hline
        \multirow{3}{1.5cm}{Perceptron\newline}     & 4096*8    &            &              \\
        \multirow{2}{0.9cm}{Weights}                 & 2048*x    & 5 bits     & 164480 bits  \\
                                                    & 128*1     &            &              \\
    \hline
        Prefetch Table\footnotemark[1]             & 1024      & 84 bits    & 86016 bits   \\
    \hline
        Reject Table\footnotemark[2]               & 1024      & 83 bits    & 84992 bits   \\
    \hline
        \multirow{4}{1cm}{Global\newline\newline}   & \multirow{4}{0.2cm}{8} & Signature (12 bits)  & \multirow{4}{1.1cm}{264 bits} \\
        \multirow{3}{1.1cm}{History\newline}        &                        & Confidence (8 bits)  &                               \\
        \multirow{2}{1.2cm}{Register}               &                        & Last Offset (6 bits) &                               \\
                                                    &                        & Delta (7 bits)       &                               \\
    \hline
        Accuracy        & 1     & C$_{total}$       & 10 bits   \\
        Counters        & 1     & C$_{useful}$      & 10 bits   \\
    \hline
        \multirow{3}{1.5cm}{Global PC\newline}      &       & $PC_1$ (12 bits)      &           \\
        \multirow{2}{1.5cm}{~Trackers}              & 3     & $PC_2$ (12 bits)      & 36 bits   \\
                                                    &       & $PC_3$ (12 bits)      &           \\
    \hline
        \multicolumn{4}{|c|}{\textbf{Total: 371392 bits = 45.33 KB}}\\
    \hline
    \end{tabular}
    \caption{SPP-Perc Storage Overhead}
    \label{tab:PPF_overhead}
\end{adjustwidth}
\end{table}

\footnotetext[1]{Components of Prefetch Table can be found in Table \ref{tab:PTable_metadata}}
\footnotetext[2]{The Reject Table does not need to maintain the useful bit as that only applies for prefetches that ultimately made through}

\begin{figure}[h]
\includegraphics[width=\columnwidth]{SPEC2017}
\caption{SPEC 2017 Single-Core IPC Speedup}
\label{Fig:SPEC2017_1core}
\end{figure}

### 2018-12-01 results.tex
Separating arch and implementation. Filter-Table and Throttler-Filter renaming
Updated results section & added graphs

Figure \_\_ shows the IPC speedup on the memory intensive subset of SPEC 2017. 
All the results have been normalized to baseline \textit{i.e.}, no
prefetching.  As can be seen, PPF yields a speedup of XX\% over
the baseline.  This is equivalent to XX\% over DA-AMPM, xx\% over BOP
and XX\% over SPP.  Out of the XX benchmarks, PPF nearly matches
or outperforms most of the prefetchers on XX traces.

At its peak, PPF manages to get the IPC speedup of a factor of
over \textbf{XX} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{XX\%} over the next best
prefetcher - SPP.  In general, benchmarks \textit{602.gcc\_s,
  603.bwaves\_s, 605.mcf\_s, 621.wrf\_s} and \textit{649.fotonik3d\_s}
benefit the most from PPF, as compared to SPP only prefetching.

On the full SPEC 2017 suite, PPF improves the baseline by XX\%,
which is YY\% more than just SPP.

<Discuss the extreme behaviour of \textit{623.xalancbmk\_s}:
Improvement over SPP is as high as 31\% and even reduces to -1.5\% on
traces of this benchmark>

<Discuss about coverage on all traces>

<Discuss cases where coverage gets specially increased>

<Discuss accuracy on all traces>

<Discuss about compelling increase in accuracy - details of
traces. Plot linear separability>

<Any case where accuracy got reduced - go into the trace. See if
linear separability is the issue. Other possible issues?>

@@

\textit{4-core environment} Figure \_\_ shows a comparison of speedups
obtained on 4-core mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of IPC obtained with running PPF.
As can be seen,
PPF offers an IPC improvement by 51.7\% on these traces.  This is
11.9\% over the next best prefetcher SPP; and XX\% and YY\% over BOP
and DA-AMPM respectively.

On a different set of fully random SPEC 2017 4-core mixes, PPF provides an IPC speedup of 
26.2\% over baseline, which is 5.7\% more than just SPP.

\textit{8-core environment} The sorted comparison of speedup obtained 
on the memory intensive 8-core mixes, is depicted in Fig \_\_.
PPF improves baseline performance by 38.6\% and that is 10.7\% 
more than using just SPP.

For completely random SPEC 2017 mixes, PPF improves performance by 
23.6\% over the baseline, corresponding to 4.8\% over SPP.

This increased improvement by PPF on a multi-core machines is not
surprising.  Since PPF has an extremely intelligent throttler, it ends up 
eliminating useless prefetches before they can create cache
pollution. This further amplifies the edge that SPP had over BOP in multi-core environment.

\textit{<See if we can use
speed-up over BOP as a measure here for showing scalability>}

@@

\subsection{Cross Validation}
\label{Results-CrossVal}
Here we try to demonstrate the robustness of PPF by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig \_\_ shows the speed-up achieved by BOP, SPP and PPF on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, PPF provides a speed-up of
44.4\% over baseline, on the memory intensive subset of SPEC 2006
benchmark.  This corresponds to 6.84\% over SPP and XX\% over BOP.  On
the whole of the SPEC 2006 suite, the speedup obtained is 22.4\%,
which is 3.34\% ahead of SPP.

For 4-core memory intensive mixes, PPF improves the baseline by 59.2\%, 
8.7\% ahead of SPP. For 8-core memory intensive mixes, the IPC speedup 
over baseline is 47.9\%, 11.5\% ahead of SPP.

As can be seen, the performance is very consistent given no prior
parameter tuning was done keeping SPEC 2006 in mind.  We attribute
this to the inherent adaptability of the perceptron model.  In
general, perceptron weights are able to adjust in real-time so as to
find the best possible correlation between the output and the
features.

>>>>

Figure~\ref{Fig:SPEC2017_1core} shows the IPC speedup obtained by
PPF, compred to BOP, DA-AMPM and SPP. 
In the figure we have depicted selected traces, where PPF shows a 
noticeable improvement or shows some degradation, along with the 
geometric mean improvement on the memory intensive subset of SPEC 2017, 
and finally for the complete SPEC 2017 benchmark. 
All the results have been normalized to baseline \textit{i.e.}, no
prefetching.

As can be seen, PPF yields a speedup of \textbf{42.8\%} over
the baseline. This is equivalent to \textbf{7.8\%} over DA-AMPM, \textbf{9.7\%} over BOP
and \textbf{6.92\%} over SPP.  Out of the 95 simpoints developed for SPEC 2017, 
PPF nearly matches or outperforms most of the prefetchers on 91 simpoints.

Only in 607.cactuBSSN and 654.roms, PPF fails to match the improvement 
offered by BOP.

At its peak, PPF manages to get the IPC speedup of a factor of
over \textbf{3.39x} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{60\%} over the next best
prefetcher -- BOP. 
In general, benchmarks \textit{603.bwaves\_s, 605.mcf\_s, 623.xalancbmk\_s} 
and \textit{649.fotonik3d\_s} benefit the most from PPF, with the speedup over
SPP ranging from \textbf{10\% to 25\%}.

On the full SPEC 2017 suite, PPF improves the baseline by \textbf{20.9\%},
which is \textbf{3.44\%} more than the next best prefetcher -- SPP.

<Discuss the extreme behaviour of \textit{623.xalancbmk\_s}:
Improvement over SPP is as high as 31\% and even reduces to -1.5\% on
traces of this benchmark>\newline

\begin{figure}[h]
\includegraphics[width=\columnwidth]{Coverage}
\caption{Fraction of Cache Misses Covered}
\label{Fig:Coverage}
\end{figure}

\textbf{Coverage}\newline
A prefetcher coverage can be defined as the ratio of number of misses 
avoided through prefetching to the number of misses in presence of no 
prefetchig.
Figure~\ref{Fig:Coverage} shows the fraction of misses in L2 and Last
Level Caches avoided by the various prefetchers.
PPF has the highest coverage of all the prefetchers simulated. Using PPF 
on SPEC 2017 benchmark, 
misses can be reduced by \textbf{62.5\%} and \textbf{82.8\%} in L2 Cache and Last Level 
Cache respectively. For the same benchmark, the next best prefetcher -- 
DA-AMPM covers \textbf{48.6\%} and \textbf{72.8\%} of the misses respectively.

This superior coverage of PPF can be attributed to aggressive 
re-tuning of the underlying SPP. At the same time, the Perceptron Filter
makes sure than this high coverage does not lead to an increased 
cache pollution.

<Discuss cases where coverage gets specially increased>
<Discuss accuracy on all traces>

\begin{figure}[h]
%%\includegraphics[width=\columnwidth]{4Core_SPEC2017}
\caption{Normalized speedup for 2 sets of 4-core workloads}
\label{Fig:4Core_SPEC2017}
\end{figure}

@@

\textit{4-core environment} Figure~\ref{Fig:4Core_SPEC2017} (a) 
shows a comparison of speedups
obtained on 4-core mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of IPC obtained with running PPF.
As can be seen,
PPF offers an IPC improvement by \textbf{51.7\%} on these traces.  This is
\textbf{XX\%} over the next best prefetcher -- \_\_; and \textbf{XX\%} 
and \textbf{YY\%} over \_\_ and \_\_ respectively.

On a different set of fully random SPEC 2017 4-core mixes, 
PPF provides an IPC speedup of 
\textbf{26.2}\% over baseline, which is \textbf{5.7}\% more than just SPP.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{8Core_SPEC2017}
\caption{Normalized speedup for 2 sets of 8-core workloads}
\label{Fig:8Core_SPEC2017}
\end{figure}

\textit{8-core environment} The sorted comparison of speedup obtained 
on the memory intensive 8-core mixes is depicted in Fig~\ref{Fig:8Core_SPEC2017}.
PPF improves baseline performance by \textbf{38.6\%} and that is \textbf{10.7\%} 
more than using just SPP. 
For completely random set of SPEC 2017 mixes, PPF improves performance by 
\textbf{23.6\%} over the baseline, corresponding to \textbf{4.8\%} over SPP.
This increased improvement achieved by PPF over the base engine SPP,
in a multi-core environment is not surprising. 
Since PPF has an extremely intelligent filter, it ends up 
eliminating useless prefetches before they can create cache
pollution in the shared LLC.

As can be seen, BOP offer a better improvement than 
SPP, for the memory intemsive mix, This can be attributed 
to BOP's inherent aggressive nature. DA-AMPM is also ahead of
SPP in both the mixes. Interestingly, in all these cases, 
PPF consistently outperforms the best performing prefetcher.

%[EB] How to justify only a small margin over DA-AMPM?

@@

\begin{figure}[ht]
\includegraphics[width=\columnwidth]{SPEC2006}
\caption{SPEC 2006 Single-Core IPC Speedup}
\label{Fig:SPEC2006_1core}
\end{figure}

\subsection{Cross Validation}
\label{Results-CrossVal}

Here we try to demonstrate the robustness of PPF by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig~\ref{Fig:SPEC2006_1core} shows the speed-up achieved by BOP, SPP and PPF on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, PPF provides a speed-up of
46.1\% over baseline, on the memory intensive subset of SPEC 2006
benchmark. This corresponds to \textbf{6.66\%} over SPP and {9.86\%} over BOP and DA-AMPM.  
On the whole of the SPEC 2006 suite, the speedup obtained is \textbf{22.4\%},
which is \text{3.34\%} ahead of SPP.

For 4-core memory intensive mixes, PPF improves the baseline by \textbf{59.2\%}, 
\textbf{8.7\%} ahead of SPP. For 8-core memory intensive mixes, the IPC speedup 
over baseline is \textbf{47.9\%}, \textbf{11.5\%} ahead of SPP.

As can be seen, the performance is very consistent given no prior
parameter tuning was done keeping SPEC 2006 in mind.  We attribute
this to the inherent adaptability of the perceptron model.  In
general, perceptron weights are able to adjust in real-time so as to
find the best possible correlation between the output and the given set of 
features.
